---
title: Logistic Regression with Gradient Descent
author: Imran Kocabiyik
date: '2018-05-08'
---

## 1. The Objective

I am implementing the learning algorithms that I learned in <a href="https://click.linksynergy.com/fs-bin/click?id=GROwIP9ejy4&offerid=467035.248&type=3&subid=0">Andrew Ng’s famous Machine Learning course</a><img border="0" width="1" alt="" height="1" src="https://ad.linksynergy.com/fs-bin/show?id=GROwIP9ejy4&bids=467035.248&type=3&subid=0" > (Affiliate Link).

I will solve a **binary classification problem** with Gradient Descent. My main objective is getting a better understanding of Gradient Descent. It will solve not only this problem, but also more complex ones like Neural Networks. I wanted to use that algorithm because it is like a Swiss Knife when solving most of the machine learning problems.  

Having this objective, I need to find a problem where the output is discrete valued. $y\in\{0,1\}$. So, I found [Titanic: Machine Learning from Disaster](https://www.kaggle.com/c/titanic) competition on Kaggle would be a good practice. It is about predicting the survival chance of the passangers.  

The below implementation is achieving 77% performance but I didn't spend too much effort to achieve higher score.  

## 2. Logistic Regression
Introducing some terms:  

$h_\theta(x) = \theta_0x_1 +\theta_1x_2+\theta_2x_3+...  = \theta^Tx$ 

Sigmoid function: $g(z) = \frac{1}{1+e^{-z}}$  

The predictions: $g(\theta^Tx)$  

What is sigmoid function? It maps a real number to a value between 0 and 1.  
It is also called Logistic function. In neural networks, it is called activation function. It can be visualized as below:  

```{r sigmoid-function, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

sigmoid <- function(x) {1/(1+exp(1)^(-x))}

data_frame(x = c(-5, 5),
           y = c(0, 1)) %>%
  ggplot(aes(x = x, y = y))+
  stat_function(fun = sigmoid)
```

$h_\theta(x) = 0.7$ can be interepreted as: x is *positive* with the probability of 0.7.  

## 3. Logistic Regression with Gradient Descent

We need a measure for divergence of the fit and we will minimize it. It is called *cost*.  

The cost function in this examples is based on [Cross Entropy Loss](https://ml-cheatsheet.readthedocs.io/en/latest/loss_functions.html#cross-entropy).  

### 3.1 Cost Function

$$
\begin{equation}
  h_\theta(x)=\begin{cases}
    -log(h_\theta(x) \text{ if } y = 1 \\
    -log(1-h_\theta(x) \text{ if } y = 0)
  \end{cases}
\end{equation}
$$

In more compact way: 


$$h_\theta(x) = -y*log(h_\theta(x))-(1-y)*log(1-h_\theta(x))$$

And cost function will be:
$$J(\theta) = -\frac{1}{m}\sum\limits_{i=1}^m[(y^{(i)})log(h_\theta(x^{(i)}))+(1-y^{(i)})log(1-h_\theta(x^{(i)}))]$$ 

### 3.2 Gradient Descent
We will minimize cost with iteration. In each iteration, we will update parameters so that cost will get lower.  This is actually _the process of learning_.  

Implementation:   

_Repeat_
$\big\{$  
$\theta_j :=\theta_j-\alpha*\frac{\partial}{\partial\theta_j}*J(\theta)$  
$\big\}$

When we do the calculus:  
_Repeat_
$\big\{$  
$\theta_j :=\theta_j-\frac{\alpha}{m}\sum\limits_{i=1}^{m}(h_\theta(x^{(i)}-y^{(i)}x_j^{(i)})$  
$\big\}$

Note that this is identical to linear regression.  

After stating some notations, now let's start the implementation:  

## 4. Implementation

### 4.1 Import Data
```{r, message = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
df_train <- read_csv('train.csv') %>% mutate(dataset = "train")
df_test <- read_csv('test.csv') %>% mutate(dataset = "test")
```

### 4.2 Data Processing
Process test and train datasets together:  

Steps: 

- Bind rows of train and test sets  
- Convert categorical columns (`Sex`, `Embarked`) to intergers.  
- Separate datasets into train and test again.

```{r, message = FALSE, message=FALSE, warning=FALSE}
# merge to datasets for data transformation
df_all <- bind_rows(df_train, df_test)

df_all <- df_all %>%  mutate(Sex = if_else(Sex =="male", true = 1L, false = 0L))

df_all <- df_all %>%
                mutate(Embarked = case_when(Embarked == "S" ~ 1L,
                                            Embarked == "C" ~ 2L,
                                            Embarked == "Q" ~ 3L,
                                            is.na(Embarked) ~ 4L,
                                            TRUE ~ 5L))

# separate datasets
df_train <- df_all %>% filter(dataset == "train")
df_test <- df_all %>% filter(dataset == "test")
```

### 4.3 Define Functions

Sigmoid Function:  
```{r, message = FALSE, message=FALSE, warning=FALSE}
# define sigmoid function
sigmoid <- function(x) {1/(1 + exp(-x))}
```

Cost Function:  
```{r, message = FALSE, message=FALSE, warning=FALSE}
# define cost function
cost <- function(x, y, theta){
  m <- nrow(x)
  hx <- sigmoid(x %*% theta)
  (1/m)*(((-t(y)%*%log(hx))-t(1-y)%*%log(1-hx)))
}
```

Gradient Function:  
```{r, message = FALSE, message=FALSE, warning=FALSE}
# gradient
grad <- function(x, y, theta){
  m <- nrow(x)
  hx <- sigmoid(x %*% theta)
  (1/m)*(t(x)%*%(hx - y))
}
```

Once we find the parameters, we will make predictions with test data.  
So, prediction Function will be defined as:  
```{r, message = FALSE, message=FALSE, warning=FALSE}
# prediction
logit_predict <- function(df, theta, features) {
  x <- as.matrix(df[,features])
  x <- cbind(1, x)
  sigmoid(x %*% theta) %>%
    as_data_frame()
}
```


### 4.4 Runing Gradient Descent
This will return the parameters which minimizes the cost.  
```{r, message = FALSE, message=FALSE, warning=FALSE}
logit_gradient_descent <- function(df, iter = 10000, learning_rate, features, response){
  
  # initial theta values: set all of them to 1
  n_independent = length(features)
  theta = matrix(rep(1, n_independent+1), nrow = n_independent+1)
  m = nrow(df)
  
  # dependent and independent variables 
  y = as.matrix(df[response])
  x <- as.matrix(df[,features])
  x <- cbind(1, x)
  
  cost_values = NA
  # run gradient descent
  for (i in 1:iter){
    theta <- theta-learning_rate*grad(x, y, theta)
    
    cost_values[i] <- cost(x, y, theta)[1,1]
    if(i > 1){
      if(cost_values[i] > cost_values[i-1]) {
        warning("Cost is not decreasing in each iteration. Lower the learning_rate")
        break
      } 
    }
  }
  theta
}
```

### 4.5 Make Predictions
```{r, message = FALSE, message=FALSE, warning=FALSE}
theta <- logit_gradient_descent(df = df_train, iter = 100000, learning_rate = 0.001,
                                features = c("Sex", "Pclass", "SibSp", "Embarked"), response = 'Survived')

predictions_train <- logit_predict(df_train, theta, features = c("Sex", "Pclass", "SibSp", "Embarked"))
df_train$predicted <- predictions_train$Survived
df_train %>% mutate(predicted = as.integer(round(predicted))) %>% 
  mutate(is_prediction_true = Survived == predicted) %>% 
  group_by(is_prediction_true) %>% 
  count()

predictions <- logit_predict(df_test, theta, features = c("Sex", "Pclass", "SibSp", "Embarked"))
```

## References
- [Coursera Machine Learning Course](https://www.coursera.org/learn/machine-learning) by Andrew Ng  